Extended Data

### 2022-03-09 doc/manuscript/Dow_manuscript_Nature.Rmd
remove new added text in discussion
#122

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. r tree_rings`).

<!--new added text-->
We find statistically, but not biologically, significant relationships between total growth and spring temperature at both sites in our dendrometer band analysis (`schematic_summary`). 
However, the slope of the relationships are in opposite directions, with warming causing decreasing stem growth at Harvard Forest but increasing stem growth at SCBI (`EDT_phenology_parameters`). 
We believe this difference is due to severe drought conditions in 1999 which obviously caused abnormal growth patterns in that year (`r EDF_pheno_Tmax_climwin`). 
Therefore, we suggest that our two sites represent two scenarios: 1. A forest which had no significant summer drought, but was only influenced by warming temperatures (SCBI). 2. A forest which experienced spring warming along with severe summer drought (Harvard Forest).

At SCBI, we see a very small increase in total stem growth (on average +0.05mm/$^\circ$C), suggesting that even in the absence of drought or inclement climate conditions during the growing season, spring warming does not enhance stem growth in a meaningful way.
At Harvard Forest, we see a decrease in total growth with warmer temperatures in the CTW (-0.15mm/$^\circ$C on average). We speculate that this result is likely a product of reductions in total stem growth related to drought effects in 1999. 
Due to the method of selection for CTW, years with significant departures from the norm can influence the analysis to a greater extent. 
Stem growth in 1999 was clearly impacted by climate conditions (`r EDF_pheno_Tmax_climwin`), and so the CTW selected by climwin maximizes the effect of these impacts (most likely the effects of drought) on stem growth. 
To examine this further, we reran our dendrometer band analysis using the month of May as the CTW. 
When we change the CTW to the month of May, we see results much closer to those found at SCBI, with a similar advancement in growth timing but a much less pronounced effect of $T_{max}$ on total growth (-0.015mm/$^\circ$C on average, a tenfold decrease compared to the CTW optimized around the drought year;`EDF_pheno_Tmax_AprilMay`). 
Once again see that there was no meaningful influence of spring temperature on total growth. 
In fact, the magnitude of the effect of the summer drought compared to the effects of spring $T_{max}$ illustrates the point that conditions during the peak growing season are much more influential to stem growth than conditions during the Spring. 
<!--end new added text-->

Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF. Maxrate_DOY for RP at HF was actually significantly delayed. This could be a consequence of the severe drought in 1999 which we've included in our analysis. My brain is completely fried to the point that I barely remember what the literature claims this max rate is related to (water stress, right?). So Maybe you're in a better position to add some discussion about this here. -->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

>>>>

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise fail to translate extended growing seasons into biologically significant increases in stem growth (Fig. `r schematic_summary`), and thereby have negligible effect on total annual growth for most species and locations (Fig. r tree_rings`).
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF. Maxrate_DOY for RP at HF was actually significantly delayed. This could be a consequence of the severe drought in 1999 which we've included in our analysis. My brain is completely fried to the point that I barely remember what the literature claims this max rate is related to (water stress, right?). So Maybe you're in a better position to add some discussion about this here. -->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
*However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.* <!--issue #122-->
